---
title: "Reproducible Research Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

<br>

##### Prepared by *TG*, December 2017

<br>

### I. Introduction

<br>

This is the first report for assesment of *Reproducible Reserach* course in the *Coursera & Johns Hopkins Data Science Specialization* . The questions assigned will be answered in each of the numebered questions below. For this asignement:

* Data comes from a personal activity monitoring device that collects data at 5 minute intervals through out the day

* The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and includes the number of steps taken in 5 minute intervals each day

* Data was taken from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

* The variables included in this dataset are:

      + **steps**: Number of steps taken in a 5-minute interval (missing values are coded as `NA`)
      + **date**: The date on which the measurement was taken in YYYY-MM-DD
      + **interval**: Identifier for the 5-minute interval in which measurement was taken
  
<br>

* The dataset is stored in a comma-separated-value (CSV) file and has a total of 17,568 observations

* Asignement questions are answered following the order given in the instructions, with a brief decsrption of the steps undertaken and followed by the code used and its output

<br>

### II. Questions

<br>

#### Q1. Loading and preprocessing the data
<br>
1. The data was loaded ensuring character variables are read as characters and not factors so they can be esaily converted to Date format, and were assigned to the object *data*. Note the code assumes the unziped data file is in the current working directory.


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

```{r, message=FALSE, warning=FALSE}

data <- read.csv("activity.csv", stringsAsFactors = FALSE)

```

<br>

2. A new factor variable *day* is created and *date* is transformed to date class to facilitate the analysis
<br>
```{r, message=FALSE, warning=FALSE}
data$day <- as.factor(data$date)
data$date <- as.Date(data$date)
str(data)

```

<br>

#### Q2. What is the mean total number of steps taken per day?
<br>
1. To abtain the total numer of steps taken each day without missing values, data is filetered with the *dplyr* package to work only with the complete cases and then summarised by *day* with the following code:

```{r, message=FALSE, warning=FALSE}

library("dplyr", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
data$complete <- complete.cases(data)
data1 <- filter(data, complete==TRUE)
by_day <- group_by(data1, day)
totday <- summarise(by_day, sum(steps))
totday

```

<br>

The histogram for the total number of steps can then be created with the *ggplot2* package:

```{r, message=FALSE, warning=FALSE}

library("ggplot2", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")

ggplot( totday, aes(x=`sum(steps)`) ) + geom_histogram( fill = "red4" ) +
      ggtitle ( "Total Steps per Day" ) + labs( x= "Steps" )

```

<br>  


2. The Mean and the median number of steps taken by day can be calculated from the dataset without mising values with

```{r}

mmday <- summarise(by_day, mean(steps), median(steps))
mmday

```

<br>

#### Q3. What is the average daily activity pattern?
<br>
1. To obtain and plot the average number of steps for each 5-minute interval without the missing values, we use the filterd dataset to summarize the data by the `mean()` of *interval*, and then plot it in a histogram

```{r}
by_interval <- group_by(data1, interval)
intavg <- summarise(by_interval, mean(steps))
names(intavg) <- c("interval", "average_steps")

ggplot(intavg, aes(y = average_steps, x = interval) ) + geom_line(color = "red4") +
      labs(title = "Average number of Steps by Interval", y = "Average Steps", x = "Interval")
```

<br>

2. The highest daily average number of steps is

```{r, message=FALSE, warning=FALSE}
intavg$interval[ which.max(intavg$average_steps)]
```

which has a value of `r round( intavg[ which.max(intavg$average_steps), 2 ], 2)` minutes.

<br>


#### Q4. Imputing missing values
<br>
1. The number of missing values in the dataset is 

```{r}
sum( is.na( data$steps ) )
```

which corresponds to roughly `r round(2304/17568*100, 2)`% of the total nuber of observations.

2. To deal with the bias in many calculations missing values may cause, an `NA` in a particular interval will be replaced by the average number of steps for the whole sample in that interval. 

<br>

3. The new database is stored in the object *data2* that replicates the original data but adds the correction. These two steps are achieved by the following code

```{r}
data2 <-data
data2$steps[is.na(data2$steps)] <- intavg$average_steps[ match(data2$interval[ is.na(data2$steps) ], intavg$interval)]
head(data2)
```

<br>

4. The impact of the correction can be seen by calculating the **total number of steps** taken by day for *data2* and plotting a historgram. A summary for the total number of steps with and without the correction is also presented afterwards to facilitate comparisons:

```{r}

by_day <- group_by(data2, day)
totday1 <- summarise(by_day, sum(steps))

ggplot( totday1, aes(x=`sum(steps)`) ) + geom_histogram( fill = "aquamarine4" ) +ggtitle ( "Total Steps per Day after Correction" ) + labs( x= "Steps" )

summary(totday1$`sum(steps)`)
summary(totday$`sum(steps)`)

```


The **mean** and **median** number of daily steps are calculetd first for the corrected dataset, and, to facilitate comparison, then for the original dataset

```{r}

by_day1 <- group_by(data2, day)
mmday1 <- summarise(by_day1, mean(steps), median(steps))
mmday1 

mmday
```

The tables and histogram show

* The average and mean didn't change significantly because the missing values were concentrated in specific days and were replaced by values that 'replicate' the structure of the data.

* Correcting the missing values added `r count(mmday1)-count(mmday)` more days of data

* This changed the median of total steps by 1 and the interquantile range, but not the mean or the minimum and maximum values

* The original mean and median number of daily steps were not altered, but the new data has now 8 more observations

<br>

#### Q5. Are there differences in activity patterns between weekdays and weekends?
<br>
1. To answer this question we have to determine first the corresponding day of the week for each date, and then use it to create a factor variable *wkday* to separate between weekdays and weekends

```{r, message=FALSE, warning=FALSE}
data2$weekday <- as.factor( weekdays(data2$date) )
weekend <- c ("Saturday", "Sunday")
data2$wkday <- as.factor( ifelse( data2$weekday %in% weekend, "weekend", "weekday") )

str(data2)

```

<br>

2. The differences in average steps can be seen with a time series plot for each type of weekday  

```{r, message=FALSE, warning=FALSE}

by_interval2 <- data2 %>% group_by(interval, wkday)
intavg2 <- by_interval2 %>% summarise(mean(steps))

names(intavg2) <- c("interval", "wkday", "average_steps")

ggplot(intavg2, aes(y = average_steps, x = interval) ) + geom_line(color="blue4") + facet_grid(wkday~.) + 
      labs(title = "Average number of Steps by Weekday", y = "Average Steps", x = "Interval")

```

